The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2020-02-14 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small corrections (typos)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2979 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e08cb5f9d03cf42da2d8fa5818c7964441a08742
final touch-ups; removed fixmes and highlighting of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2981 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0178b6bb6aea3308d524bc0fafb3f4e4e2306cfb

\vspace*{0.4cm}

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with strong variation 
and complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While almost all algorithms proposed in the literature
follow the general idea of splitting the data such that some objective function is optimized
locally, they differ in their specific approaches to selecting a split variable and the 
corresponding split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separate second step is a more 
promising strategy. \hlc{The most popular algorithms adopting such an unbiased recursive
partitioning strategy are} QUEST \citep{Loh+Shih:1997},
%% CHAID \citep{Kass:1980},
GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}. \hlc{QUEST and CTree have been suggested as
nonparametric decision trees initially and have acquired more than 1,300 and 2,400 citations
on Google Scholar, respectively. GUIDE and MOB have been designed for general model-based
trees and have both received more than 400 citations.
All of these share the same basic concept of selecting that covariate for splitting which shows the
highest association to the response variable as captured by the $p$-value in the corresponding
statistical test.} While QUEST
%%, CHAID,
and its successor GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on central limit theorems
for the parameter estimators. 
All these approaches have been shown to work well for various situations, however, the relative
(dis)advantages of the testing strategies have not yet been investigated and compared in detail.

\fixme{``Why CTree, GUIDE, and MOB?''}

Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. 
In particular, the approach from the GUIDE algorithm \hlc{(which essentially encompasses QUEST)}
is compared to the one of the CTree and MOB algorithms by investigating the building blocks of 
their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Dichotomization of these residuals or scores.
(3)~Categorization of possible split variables.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

The remainder of this manuscript is structured as follows:
Section~\ref{sec:urp} reviews unbiased tree models, starting with the general algorithmic idea
(Section~\ref{sec:generic_alg}) followed by the specific algorithms CTree (Section~\ref{sec:ctree}), 
MOB (\ref{sec:mob}), and GUIDE (\ref{sec:guide}), before possible pruning strategies 
(\ref{sec:pruning}) are discussed.
In Section~\ref{sec:unify} the unifying framework for testing strategies in unbiased model-based
recursive partitioning algorithms is presented.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are illustrated and discussed in Section~\ref{sec:results}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and each possible transformed split variable $g(Z_j)$, $j = 1, \dots, J$.
The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
this encompasses ranks, scores, indicator functions, etc.\ and can also be multidimensional.
For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can capture location and scale of $Y$ via
$h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fitted to the response $Y$ with some covariate(s)~$X$ employed as 
\textit{regressor} variable(s),
then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
the residuals in a linear model with regression coefficients $\beta$. Moreover,
$s$ can be the score function pertaining to the objective function (or loss function)
$\ell$:
$$
s(Y, X, \beta) = \frac{\partial \ell(Y, X, \beta)}{\partial \beta}
$$
The estimate of the model parameters $\hat{\beta}$ is obtained by optimizing
the sum of the objective function $\ell$, aggregated over all observations.
This framework includes many different M-type estimators as special cases, including maximum 
likelihood and ordinary least squares estimation. For a $K$-dimensional parameter
$\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
is also a $K$-dimensional vector, i.e., the gradient contribution of the $i$-th observation.
Thus, the $N \times K$-matrix consisting of these scores or gradient contributions for all 
$i = 1, \dots, N$ observations is a natural candidate for the observation-wise goodness-of-fit
measure as described in the previous Section~\ref{sec:generic_alg}.

@@

If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a Pearson 
correlation test. For one numeric and one categorical variable essentially a one-way ANOVA
(analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
Thus, in the general CTree framework many types of tests can be specified by selecting suitable
transformations $g$ and $h$. While originally
conceived for nonparametric models, it is easy to adapt CTree to model-based testing and 
recursive partitioning by choosing a model-based $h$ transformation as argued above
\citep[see the concrete examples in][]{Zeileis+Hothorn:2013,Seibold+Zeileis+Hothorn:2016}.

@@

In contrast to CTree the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was explicitly
designed for a model-based goodness-of-fit measure in order to embed parametric models
into a regression tree framework. Thus, MOB is based on an objective/loss function $\ell$
and corresponding score function $s$. The original paper considered generalized linear models
(GLMs) and survival regression models but subsequently various other models have been applied as well,
including beta regression \citep{Gruen+Kosmidis+Zeileis:2012}, psychometric item response theory
models \citep{Strobl+Kopf+Zeileis:2015}, or mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
But just like CTree can be applied to parametric models, MOB conversely also encompasses
simple regression and classification trees, e.g., by choosing an intercept-only model.

@@

To construct a statistical test two additional transformations are carried out:
(1)~The residuals $r(Y,X,\hat{\beta})$ are dichotomized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the dichotomized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test. More details on the
applied test statistic can be found in Appendix~\ref{app:teststat_guide}.

\subsection{Selecting a split point}
\fixme{``Why is the choice of the split point not part of this paper?''}

\hlc{Once a split variable has been selected, the selection of the corresponding
split point is less controversial. The most natural strategy is to use an exhaustive
search over all possible split points and then select the one that optimizes the
objective function (or loss function). In fact, this is the strategy adopted by both
model-based approaches, GUIDE and MOB. Only CTree uses a slightly different approach,
based on maximizing a two-sample test statistic, because the algorithm as originally
proposed did not employ an explicit objective function. However, it can be shown
that in some settings the two approaches coincide} \citep{Zeileis+Hothorn:2013}
\hlc{while the differ somewhat more generally. As we focus on model-based partitioning
here we employ the exhaustive model-based search for split point selection in
all three algorithms considered.}

@@

\fixme{``Do the building blocks cover all possible options?''}

Note that there are further differences in the testing strategies between the three algorithms, e.g.,
using conditional vs.\ unconditional approximations of the null distributions. However, this difference
has relatively little influence compared to the other building blocks considered in detail here.
Moreover, both similarities and relative differences between these approaches have been previously
discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
\hlc{Therefore, this aspect is not considered again here but focus is given to the building
blocks selected above.}

@@

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
can depend on the true split variable $Z_1$. More specifically, three different variations
are considered for the coefficients $\beta_0$ and $\beta_1$:
%
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
(at $0$).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
See Figure~\ref{fig:dgp_stump} for an illustration. More precisely, the coefficient that
changes~($\beta_{k-1}$ with $k=1,2=K$) switches between two values at the split point $\xi$:
%
\begin{align*}
\beta_{k-1}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{k-1}  \quad \text{if } Z_1 \leq \xi \\
+\delta \cdot (-1)^{k-1}  \quad \text{if } Z_1 > \xi
\end{cases}
\end{align*}
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.

@@

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, changetype= "abrupt", 
               variation = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three variations: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

@@

\begin{figure}[p!]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5.2, height=4.4>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure
applied in the data generating process for the ``tree'' scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

@@

>>>>

%\vspace*{0.4cm}

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a given data set can be very challenging, 
especially if the data contains lots of different features with strong variation 
and complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on a set of covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation can be 
done by applying a tree algorithm. While almost all algorithms proposed in the literature
follow the general idea of splitting the data such that some objective function is optimized
locally, they differ in their specific approaches to selecting a split variable and the 
corresponding split point. 
Some of the first tree algorithms (e.g., AID, \citealp{Morgan+Sonquist:1963}; 
CART, \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separate second step is a more 
promising strategy. The most popular algorithms adopting such an unbiased recursive
partitioning strategy are QUEST \citep{Loh+Shih:1997},
%% CHAID \citep{Kass:1980},
GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006}, 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}. QUEST and CTree have been suggested as
nonparametric decision trees initially and have acquired more than 1,300 and 2,400~citations
on Google Scholar, respectively. GUIDE and MOB have been designed for general model-based
trees and have both received more than 400 citations.
All of these share the same basic concept of selecting that covariate for splitting which shows the
highest association to the response variable as captured by the $p$-value in the corresponding
statistical test. While QUEST
%%, CHAID,
and its successor GUIDE employ 
statistical significance tests for contingency tables, CTree applies permutation tests in 
a conditional inference framework and MOB uses fluctuation tests based on central limit theorems
for the parameter estimators. 
All these approaches have been shown to work well for various situations, however, the relative
(dis)advantages of the testing strategies have not yet been investigated and compared in detail.

%\fixme{``Why CTree, GUIDE, and MOB?''}
%
Therefore, in this paper the focus will be put on that first step of tree algorithms,
i.e., the task of selecting the best split variable in a given (sub)sample. 
In particular, the approach from the GUIDE algorithm (which essentially encompasses QUEST)
is compared to the one of the CTree and MOB algorithms by investigating the building blocks of 
their testing strategies in which they differ:
(1)~Variation of the goodness-of-fit measure for the response: using residuals or full model scores.
(2)~Dichotomization of these residuals or scores.
(3)~Categorization of possible split variables.
Apart from these three main factors further aspects such as the approximation of the null
distribution (conditional vs.\ unconditional) or the type of test statistic (maximally 
selected vs.\ sum of squares) will be considered as well.
For this purpose, a unifying framework for testing strategies in unbiased model-based tree 
algorithms is presented such that each of the three strategies in GUIDE, CTree, and MOB can be 
obtained by a specific combination of the available building blocks. This allows to systematically
vary the building blocks and assess the power of the resulting inference procedure. Moreover,
it is investigated whether the performance of the inference impacts the performance of the
trees differently under pre- vs.\ post-pruning.

@@

The remainder of this manuscript is structured as follows:
Section~\ref{sec:urp} reviews unbiased tree models, starting with the general algorithmic idea~(Section~\ref{sec:generic_alg}) followed by the specific algorithms CTree~(Section~\ref{sec:ctree}), 
MOB~(\ref{sec:mob}), and GUIDE~(\ref{sec:guide}), before possible pruning strategies~(\ref{sec:pruning}) are discussed.
In Section~\ref{sec:unify} the unifying framework for testing strategies in unbiased model-based
recursive partitioning algorithms is presented.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are illustrated and discussed in Section~\ref{sec:results}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and each possible transformed split variable $g(Z_j)$, $j = 1, \dots, J$.
The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
this encompasses ranks, scores, indicator functions, etc.\ and can also be multidimensional.
For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can capture location and scale of $Y$ via
$h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fitted to the response $Y$ with some covariate(s)~$X$ employed as 
\textit{regressor} variable(s),
then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
the residuals in a linear model with regression coefficients $\beta$. Moreover,
$s$ can be the score function pertaining to the objective function (or loss function)
$\ell$:
$$
s(Y, X, \beta) = \frac{\partial \ell(Y, X, \beta)}{\partial \beta}
$$
The estimate of the model parameters $\hat{\beta}$ is obtained by optimizing
the sum of the objective function $\ell$, aggregated over all observations.
This framework includes many different M-type estimators as special cases, including maximum 
likelihood and ordinary least squares estimation. For a $K$-dimensional parameter
$\beta$ the score function evaluated for the $i$-th~observation~$s(y_i, x_i, \beta)$
is also a $K$-dimensional vector, i.e., the gradient contribution of the $i$-th~observation.
Thus, the $N \times K$-matrix consisting of these scores or gradient contributions for all 
$i = 1, \dots, N$ observations is a natural candidate for the observation-wise goodness-of-fit
measure as described in the previous Section~\ref{sec:generic_alg}.

@@

If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a \mbox{Pearson} correlation test. For one numeric and one categorical variable essentially a one-way 
ANOVA~(analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
Thus, in the general CTree framework many types of tests can be specified by selecting suitable
transformations $g$ and $h$. While originally
conceived for nonparametric models, it is easy to adapt CTree to model-based testing and 
recursive partitioning by choosing a model-based $h$ transformation as argued above
\citep[see the concrete examples in][]{Zeileis+Hothorn:2013,Seibold+Zeileis+Hothorn:2016}.

@@

In contrast to CTree the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was explicitly
designed for a model-based goodness-of-fit measure in order to embed parametric models
into a regression tree framework. Thus, MOB is based on an objective/loss function $\ell$
and corresponding score function $s$. The original paper considered generalized linear models
(GLMs) and survival regression models but subsequently various other models have been applied as well,
including beta regression \citep{Gruen+Kosmidis+Zeileis:2012}, psychometric item response theory models~\citep{Strobl+Kopf+Zeileis:2015}, or mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
But just like CTree can be applied to parametric models, MOB conversely also encompasses
simple regression and classification trees, e.g., by choosing an intercept-only model.

@@

To construct a statistical test two additional transformations are carried out:
(1)~The residuals~$r(Y,X,\hat{\beta})$ are dichotomized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the dichotomized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test. More details on the
applied test statistic can be found in Appendix~\ref{app:teststat_guide}.

\subsection{Selecting a split point}
%\fixme{``Why is the choice of the split point not part of this paper?''}

Once a split variable has been selected, the selection of the corresponding
split point is less controversial. The most natural strategy is to use an exhaustive
search over all possible split points and then select the one that optimizes the
objective function (or loss function). In fact, this is the strategy adopted by both
model-based approaches, GUIDE and MOB. Only CTree uses a slightly different approach,
based on maximizing a two-sample test statistic, because the algorithm as originally
proposed did not employ an explicit objective function. However, it can be shown
that in some settings the two approaches coincide \citep{Zeileis+Hothorn:2013}
while they differ somewhat more generally. As we focus on model-based partitioning
here we employ the exhaustive model-based search for split point selection in
all three algorithms considered.

@@

%\fixme{``Do the building blocks cover all possible options?''}
%
Note that there are further differences in the testing strategies between the three algorithms, e.g.,
using conditional vs.\ unconditional approximations of the null distributions. However, this difference
has relatively little influence compared to the other building blocks considered in detail here.
Moreover, both similarities and relative differences between these approaches have been previously
discussed, e.g., in \cite{Hothorn+Zeileis:2008} and \cite{Zeileis+Hothorn:2013}.
Therefore, this aspect is not considered again here but focus is given to the building
blocks selected above.

@@

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and/or the slope parameter $\beta_1$
can depend on the true split variable $Z_1$. More specifically, three different variations
are considered for the coefficients $\beta_0$ and $\beta_1$ (see Figure~\ref{fig:dgp_stump} for an illustration):
%
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
(at $0$).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}
%
More precisely, the coefficient that
changes~($\beta_{k-1}$~with $k=1,2=K$) switches between two values at the split point $\xi$:
%
\begin{align*}
\beta_{k-1}(Z_1) = \begin{cases}
-\delta \cdot (-1)^{k-1}  \quad \text{if } Z_1 \leq \xi \\
+\delta \cdot (-1)^{k-1}  \quad \text{if } Z_1 > \xi
\end{cases}
\end{align*}
In that way, the type of variation is the same for $\beta_0$ and $\beta_1$, however, in 
opposite directions.

@@

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, changetype= "abrupt", 
               variation = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response~$Y$ on regressor~$X$ 
with effect size $\delta = 1$ and for three variations: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

@@

\begin{figure}[p!]
\begin{center}
\setkeys{Gin}{width=1\linewidth}
\minipage{0.5\textwidth}
<<dgp_tree1, fig=TRUE, echo=FALSE, width=5.2, height=4.4>>=
par(mar=c(0,0,0,0), oma=c(0,0,0.4,0))
plot(p, edge_panel = ep, terminal_panel = tp,
     tp_args = list(FUN = identity, width = 15), 
     ip_args = list(fill = paltrees[c(1, 2)]),
     drop_terminal = TRUE, tnex = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_tree2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,4))
plot(x = d$x[d$z2<=xi], y = d$y[d$z2<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(-1,-1), col = pal_dgp[2], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[3], lwd = 2)
points(x = d$x[d$z2>xi & d$z1<=xi], y = d$y[d$z2>xi & d$z1<=xi], pch = 19, 
       col = alpha(pal_dgp[2], 0.4), cex = 0.7)
points(x = d$x[d$z2>xi & d$z1>xi], y = d$y[d$z2>xi & d$z1>xi], pch = 19, 
       col = alpha(pal_dgp[3], 0.4), cex = 0.7)
legend("top", c(expression(z[2]<=xi), 
                expression(z[2]>xi~" and "~z[1]<=xi),
                expression(z[2]>xi~" and "~z[1]>xi)), 
       col = c(pal_dgp[1], pal_dgp[2], pal_dgp[3]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.3)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = 0.1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[3], las = 1, line = 0.5, at = -0.6)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2.7)
@
\endminipage
\caption{\label{fig:dgp_tree}Left panel: True tree structure
applied in the data generating process for the ``tree'' scenario 
with the true model parameters $\beta_0$ and $\beta_1$ for each terminal panel.
Right panel:~Bivariate plot of the response $Y$ on regressor~$X$ 
with effect size $\delta = 1$.}
\end{center}
\end{figure}

@@

\section*{Competing interests statement}
The authors declare that they have no competing interests.
